\section{Script language}
\label{sec:language}

In this section we give a more detailed description of the script language.
We describe the types in Section~\ref{ssec:types}.  We describe expressions in
Section~\ref{ssec:expressions}.  We describe declarations of values and
functions in Section~\ref{ssec:declarations}.  We then describe cell write
commands in Section~\ref{ssec:cell-writes}, and |for| loop iterations in
Section~\ref{ssec:for-loops}. 

%%%%% 

\subsection{Types}
\label{ssec:types}

The script language contains the standard types |Int| of (32-bit) integers,
|Float| of (32-bit) floating point numbers, |Boolean| of booleans (with values
|true| and |false|), and |String| of strings.  

These four types (and no others) can be entered in cells of the spreadsheet.
The application decides the type of each value based on how it is entered,
mostly in the obvious way; for example, if a value looks like an integer, it
is given the type |Int|.  In most cases, a |String| can be entered by simply
typing the text, or can (optionally) be enclosed in quotation marks.  However,
in the case that a |String| is required that looks like a value of another
type, it \emph{must} be entered inside quotation marks (e.g.~|"3"|).  A
literal quotation mark can be entered as \SCALA{\\"} (e.g.~\SCALA{"He said
  \\"Hello\\""}).

The types |Row| and |Column| represent rows and columns, respectively. 

If |T| is a type, then |List[T]| represents the type of lists whose elements
are from~|T|.

If |T1| and |T2| are types, then |(T1,T2)| is the type of pairs $(x,y)$ with
$x \in \sm{T1}$ and $y \in \sm{T2}$.  This extends in the obvious way to
tuples of size up to four.

If |T1| and |T2| are types, then |T1 => T2| is the type of functions from |T1|
to~|T2|.  Note that the constructor ``|=>|'' associates to the right, so
\SCALA{T1 => T2 => T3} means \SCALA{T1 => (T2 => T3)}.



%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Expressions}
\label{ssec:expressions}

\subsubsection{Standard expressions}

The script language contains the normal arithmetic operators, |+|, |-|, |*|
and~|/|, over |Int| and |Float| (over |Int|, |/| performs rounding), and
\SCALA{\%} over |Int|.  The language also contains the normal equality (|==|),
inequality (|!=|), and order (|<|, |<=|, |>|, |>=|) operators.   Note
that |Int| and |Float| are distinct types, so can't (currently) be combined
using these operators.  Unary negation can be written using |-|; note that the
expression might need to be included in parentheses when the |-| could be
interpreted as a binary operator, for example,
\begin{scala}
  { val x = 2
    (-1)
  }
\end{scala}
%
The built-in functions |toFloat: Int => Float| and
|toInt: Float => Int| can be used to convert between these types.

The language includes the normal boolean binary operators, |&&| and
\SCALA{\|\|}.  Boolean negation is written using the function |not|.

Conditional expressions are written using C/Java/Scala-style syntax: 
\begin{scala}
  if(£\syntax{test}£) £\syntax{expression}£ else £\syntax{expression}£
\end{scala}
%
The two branches must have the same type, and this is the overall type of the
conditional expression. 

Functional applications can be written using parentheses, in the normal way,
as in Figure~\ref{fig:example-script}.  However, as in Haskell, the
parentheses can be omitted for a function of one argument, where the actual
parameter does not logically need parentheses: for example, |f 3|, or \SCALA{g
  #A3}.  Function application associates to the left, and binds tighter than
infix operators; for example, \SCALA{f g true + h \#A} represents
\SCALA{((f(g))(true)) + (h(\#A))}.

String literals are written inside double quotation marks, as in most other
languages, e.g.~|"Hello"|.  Special characters can be written using
\SCALA{\\}, for example, \SCALA{"He said \\"Hello\\".\\n"}.

Tuple expressions are written inside parentheses, using commas, for example
\SCALA{(x+y, 3.6, x>y)}.  Tuples must contain at least two, and at most four
components.  When a function is applied to a tuple expression, an extra set of
parentheses is required, for example |f((x,y))| (since the expression |f(x,y)|
would imply that |f| takes two arguments, rather than a single tuple
argument).  The functions |get1|, |get2|, |get3|, and |get4| extract the
corresponding component from a tuple; for example $\sm{get2((2,4,6,8))} =
\sm{4}$.

A \emph{block expression} consists of a sequence of declarations, followed by
an expression, separated by semicolons or newlines, and enclosed in braces;
for example,
\begin{scala}
  { def square(x: Int) = x*x; val y2 = square y; val z2 = square z; y2+z2 }
\end{scala}
The declarations are evaluated in order; then the value of the final
expression is the value of the block expression.

%%%%%

\subsubsection{List expressions}

List literals are written in square brackets, for example |[1,2,3]| or |[]|.
All the elements of the list must be of the same type.  With the empty list,
it is sometimes necessary to give the type, to help the type checker, for
example |[]: List[Int]|.

List concatenation is written using |::|, pronounced ``cons''.  This operator
takes an element on the left and a list on the right, and returns their
concatenation; for example $\sm{3 :: [4,5]} = \sm{[3,4,5]}$.  The function
|head| returns the first element of a non-empty list; for example $\sm{head
  [3,4,5]} = \sm{3}$.  The function |tail| returns the list containing all
except the first element of a non-empty list; for example $\sm{tail [3,4,5]} =
\sm{[4,5]}$.  The function |isEmpty| tests whether a list is empty. 

Lists of consecutive |Int|s can be constructed using the infix operators |to|
and |until|.  The list |m to n| produces the list from~|m| to~|n| inclusive;
for example, $\sm{3 to 5} = \sm{[3,4,5]}$.  The list |m until n| is similar,
but does not include~|n|; for example, $\sm{3 until 5} = \sm{[3,4]}$.  These
operators can also be applied to row and column values; see below. 

List comprehensions provide a way to define a list, in a similar style to
mathematical set comprehensions.  The expression \SCALA{[e \| x <- xs]}
generates the list containing the value of~|e| for each~|x| in~|xs|.  For
example,
%
\begin{eqnarray*}
\mbox{\SCALA{[x*x \| x <- [3,4,5]]}} & = & \sm{[9,16,25]}.
\end{eqnarray*}
%
A term of the form |x <- xs| is called a \emph{generator}.  A list
comprehension may have multiple generators, separated by commas; an earlier
generator is iterated over more slowly than a later generator.  For example,
\begin{eqnarray*}
\mbox{\SCALA{[x*y | x <- [2,3], y <- [4,5]]}} & = & \sm{[8,10,12,15]}.
\end{eqnarray*}
%
A list comprehension may also have one or more boolean filters; a term is
produced only for values of the variables that satisfy the filter.  For
example, 
%
\begin{eqnarray*}
\mbox{\SCALA{[x*y | x <- [2,3], y <- [4,5], x+y != 7]}} & = & \sm{[8,15]}.
\end{eqnarray*}

%%%%%

\subsubsection{Row, column, and cell expressions}

Row and column literals are written using |#|, e.g.~|#A|, |#B|, etc., and
|#0|, |#1|, etc.  A row or column value can have an |Int| added to it or
subtracted from it, with the obvious effect.  For example, $\sm{\#C} + 2 =
\sm{\#E}$, and $\sm{\#C} - 2 = \sm{\#A}$.  Arithmetic underflow, e.g.~from
$\sm{\#C} - 3$, gives an error.  Similarly, a row value can have another row
value subtracted from it; and a column value can have another column value
subtracted from it.  For example, $\sm{\#5} - \sm{\#2} = \sm{\#E} - \sm{\#B} =
3$. 

Lists of row and column values can be formed using the |to| and |until|
operators.  For example \SCALA{#B to #D} $=$ \SCALA{[#B,#C,#D]}, and \SCALA{#2
  until #4} $=$ \SCALA{[#2,#3]}.


If |c| and |r| are a |Row| and |Column|, respectively, then |Cell(c,r)| is a
\emph{cell expression}: it represents the value in the corresponding cell of
the spreadsheet.  This can be abbreviated in the case of row and column
literals; for example \SCALA{\#D3} is shorthand for \SCALA{Cell(\#D, \#3)}.

The type checker needs to be able to deduce a type, or a set of possible
types, for each cell expression.  When the script is executed, the application
checks if the corresponding cell in the spreadsheet has an expected type, and
signals an error otherwise.

One option is to provide an explicit type for a cell expression.  For example,
\SCALA{\#D3: Int} gives the type |Int| to the cell expression.

Alternatively, the type can often be deduced automatically if a function is
applied to the cell expression: this implies that its type must match the
parameter type for the function.  For example, in line~\ref{line:write-facts}
of Figure~\ref{fig:example-script}, the expression |fact(Cell(In,r))| implies
that the cell expression must have type |Int|, since |fact| takes a parameter
of type~|Int|.

Similarly, the type deduction can sometimes be made when a binary operator is
applied to a cell expression.  For example, consider the expression \SCALA{1 +
  #A3}; the type checker works from left to right, so it can deduce that the
cell expression must have type |Int|.  However, the explicit type is necessary
in the similar expression \SCALA{#A3:Int + 1}, because |+| is overloaded, and
the type checker works from left to right.

Finally, a |match| expression allows the cell to have one of several different
types, and for the value of the expression to depend on that type.  An example
was given on line~\ref{line:cell-match} of Figure~\ref{fig:example-script}.
In general, a |match| expression takes the form
\begin{scala}
  £\syntax{cell expression}£ match{ £\syntax{cases}£ } 
\end{scala}
where \syntax{cases} is a sequence of cases, separated by semicolons or
newlines.  Each case is of the form
\begin{scala}
  case £\syntax{pattern}£ => £\syntax{expression}£
\end{scala}
%
Each pattern takes one of the following forms.
\begin{itemize}
\item |Empty|, which matches an empty cell;

\item \syntax{name}|: |\syntax{type}, where \syntax{type} is one of |Int|,
  |Float|, |String|, or |Boolean|; this matches a value of the given type, and
  binds the name to that value in the subsequent expression;
 
\item |_|\,|: |\syntax{type}, which is similar to the previous form, except no
  name is bound;

\item |_|\,, which matches any type.
\end{itemize}
%
In a |match| expression, the cases are tried in the order given; the first
case that matches is selected, and the corresponding expression is evaluated.
If no case matches, an error is signalled.  Within a |match| expression, all
the expressions on the right-hand side of cases must have the same type.  


%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Declarations}
\label{ssec:declarations}

Declarations, of values and functions, may be made either at the top level of
a script, or within a local block.  Standard scoping rules apply. 

A \emph{value declaration} is written using the keyword |val|.
Figure~\ref{fig:example-script} contains several examples.

A \emph{function declaration} is written using the keyword |def|.  Again,
Figure~\ref{fig:example-script} contains several examples.  Each parameter of
the function should be given a type.  

A return type must be given for a recursive function, or if a forward
reference is made to the function: this is because the type checker considers
declarations in the order they appear, but can use claimed types of later
functions.  For example, the return type for |f| is necessary in the
following, since the use of~|f| precedes its definition:
%
\begin{scala}
val y = f(3); def f(x: Int): Int = x+1
\end{scala}
%
However, no return type is necessary in the following, since the use of~|f| is
after its definition:
%
\begin{scala}
def f(x: Int) = x+1; val y = f(3) 
\end{scala}

A function can be defined that takes several vectors of parameters.  For
example: 
%
\begin{scala}
def add(x: Int, y: Int)(z: Int) = x+y+z
val a = add(4,5) 6; val f = add(6,7); val b = f 8
\end{scala}
%
Here |f| is a function of type |Int => Int|.

Parametric polymorphism is supported in a style similar to as in Scala.  Type
parameters are introduced by enclosing them in square brackets after the name
of a function, for example:
%
\begin{scala}
def apply[A,B](f: A => B, x: A): B = f(x)
\end{scala}
%
This defines such a function for all types~|A| and~|B|.  For example
\begin{scala}
def double(x: Float) = 2*x
val a = apply(head, [1,2,3]); val b = apply(double, 3.4)
\end{scala}

Higher-order functions are supported in a style similar to Scala and Haskell. 
For example, we can define a function~|map| such that |map f| applies~|f| to
each member of a list (of the approriate type):
%
\begin{scala}
def map[A,B](f: A => B)(xs: List[A]): List[B] = 
  if(isEmpty xs) [] else f(head xs) :: map f (tail xs)
\end{scala}

Types that allow equality tests, using |==| and |!=|, are called
\emph{equality types}; examples include the atomic types, |List[A]| if |A| is
itself an equality type, but not function types.  Some polymorphic functions
can only be applied to equality types.  This is denoted by including a type
assumption such as \SCALA{A <: Eq} to specify that |A| must be an instance of
the equality type class~|Eq|.  For example, the following function removes the
first occurrence (if any) of a value~|x| from a list; but it uses an equality
test involving~|x|, so the type of~|x| must be an equality type.
%
\begin{scala}
def remove[A <: Eq](x: A)(xs: List[A]): List[A] =
  if(isEmpty xs) [] else if(head xs == x) tail xs else x :: remove x (tail xs)
\end{scala}

The language allows overloading of names of functions declared using |def|s
(but not |val|s).  For example,
\begin{scala}
def double(n: Int) = 2*n
def double(x: Float): 2*x
\end{scala}
%
When an overloaded function is applied, the type checker needs to be able to
determine which instance is intended, for example,
\begin{scala}
val m = double(3) // Int => Int instance.
\end{scala}
%
In some cases, it is necessary to give an explicit type to the argument,
particularly in the case of a cell read; for example,
\begin{scala}
val y = double(#A4: Float) // Float => Float instance.
\end{scala}
%
To allow type deductions to be made, overloaded functions must have disjoint
types for their first vectors of parameters, and may not be polymorphic. 

\subsection{Assertions}

The basic form for an assertion is |assert(|\syntax{expression}|)|, where
the expression should evaluate to a boolean.  When the script is executed, if
that boolean is|false|, an error is generated.

A generalisation is |assert(|\syntax{expression}|, |\syntax{expression}|)|,
where the second expression should evaluate to a string.  If the assertion
fails, that string is included in the error message. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\subsection{Cell writes}
\label{ssec:cell-writes}

The syntax for a cell write is \syntax{cell-expression} = \syntax{expression}.
Figure~\ref{fig:example-script} contains examples where the cell expression on
the left-hand side uses the |Cell| construction.  The \SCALA{\#} construction
can also be used on the left-hand side; for example,
\begin{scala}
#A4 = #A3:Int + 2
\end{scala}

Cell writes may not occur within a function definition, or within a block
expression: they may appear only at the top level or within a |for| statement.

Recall that the application treats any attempt to write to a non-empty cell as
an error. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%


\subsection{{\scalashape for} loops}
\label{ssec:for-loops}

The basic form of a |for| loop is
\begin{scala}
for (£\syntax{name}£ <- £\syntax{list}£) £\syntax{statements}£
\end{scala}
where \syntax{statements} is either a single statement (a declaration, cell
write, or |for| loop), or several statements enclosed in braces.  For example,
%
\begin{scala}
for(r <- #0 to #5){ 
  val x = Cell(#A, r): Int; Cell(#B, r) = 2+x; Cell(#C, r) = 2*x
}
\end{scala}

A |for| loop may have multiple generators, or boolean filters; for example,
%
\begin{scala}
for(r <- #0 to #5; c <- [#A, #C]; if r != #3) 
  Cell(c+1,r) = Cell(c,r):Int + 3
\end{scala}

Like cell writes, a |for| loop may not occur within a function definition, or
within a block expression: it may appear only at the top level or within
another |for| statement.
